Data Imputation as a Query Optimization

### 2016-12-01 paper/algorithm.tex
continue formatting algorithm section

\subsection{Imputation placement}
Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan is associated with a set of dirty attributes $D$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each base table using the table statistics, which track the number of null values in each column. If we apply an imputation operator to a relation $R$ with a dirty vector $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty vector $D' = D \ C$. Projections produce a new dirty vector by dropping each attribute that is not in the projection. A join R⨝_ψ▒R^′  ⁆ with dirty vectors D⁆ and D^′  produces a dirty vector D OR D^′⁆.  Filters do not change the dirty vector. [MS It's possible that a filter φ_t   removes all of the rows that had null-values in some column t` != t, thus removing t` from the dirty vector.]

@@

@@

# For a plan and a set of necessary attributes, add the allowed amount of imputation.
AddImpute(q, attrs):
	Dattr = Dirty(q) intersect attrs
	if Dattr = empty:
		return {q}
	else:
		return {impute_all(q), impute_min(q, Dattr), drop_min(q, Dattr)}

# For a set of plans with different dirty sets, return the best plan for each dirty set.
OptRel(Q):
	D = {Dirty(q) | q in Q}
	return { argmin_q Cost(q) s.t. Dirty(q) = d | d in D}

# Given a table and a filter predicate, return the optimal plans.
OptFilter(t⁆,φ⁆):
	return OptRel(AddImpute(t, Attrs(φ)))

# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
OptJoin(T⁆, join_preds):
	Q = empty
	for t, pred in T:
		S = T \ {t, pred}
		# Get the optimal plans for the left and right arguments to the join.
		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
			if r_l, r_r in join_preds:
				# Add imputation to the left and right arguments.
				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
	return OptRel(Q)
# Extended cost calculation to account for information loss/computation cost in imputations
Cost(q):
  case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
  case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
  case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
  case _ => normal cost calculation

@@

impute_cost(t):
  # alpha: some scaling factor
  alpha * (ntuples(t) + ct(attr(t)))

>>>>

\subsection{Imputation placement}
\label{sec:placement}
Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan is associated with a set of dirty attributes $D$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each base table using the table statistics, which track the number of null values in each column. If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \ C$.  Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$. A join $R_1 \Join_\psi R_2$ with dirty sets $D_1$ and $D_2$  produces a dirty set $D' = D_1 \cup D_2$.  Filters do not change the dirty set.

@@

The query planner must select a join ordering in addition to placing imputation operators as described in Section~\ref{sec:placement}.

@@

\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
    \Function{AddImpute}{$q, C_{req}$}
    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
    \If{$C_{min} = \emptyset$}
    \State \Return $\{q\}$
    \Else
    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
    \EndIf
    \EndFunction

    \State

    \Require{$Q$ is a set of query plans.}
    \Ensure{Returns an optimal query plan for each distinct dirty set in $Q$.}
    \Function{OptRel}{$Q$}
    \State $D \gets \{\Call{Dirty}{q} ~|~ q \in Q\}$
    \State \Return $\{\argmin_{q \in Q \land \Call{Dirty}{q} = d} \Call{Cost}{q} ~|~ d \in D\}$
    \EndFunction

    \State

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
    \Function{OptFilter}{$t, \phi$}
    \State \Return $\Call{OptRel}{\Call{AddImpute}{t, \Call{Attrs}{\phi}}}$
    \EndFunction

    \State

    \Require{$Q$ is a set of query plans and $\Psi$ relates query plans with join predicates.}
    \Function{OptJoin}{$Q, \Phi$}
    \State $Q \gets \emptyset$
    \For{$q \in Q$}

    \EndFor
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for query planning with imputations.}
\end{algorithm}

\begin{verbatim}		
# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
OptJoin(T⁆, join_preds):
	Q = empty
	for t, pred in T:
		S = T \ {t, pred}
		# Get the optimal plans for the left and right arguments to the join.
		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
			if r_l, r_r in join_preds:
				# Add imputation to the left and right arguments.
				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
	return OptRel(Q)
# Extended cost calculation to account for information loss/computation cost in imputations
Cost(q):
  case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
  case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
  case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
  case _ => normal cost calculation

@@

impute_cost(t):
  # alpha: some scaling factor
  alpha * (ntuples(t) + ct(attr(t)))
\end{verbatim}
